A Dearth of Young Stars and Planets in the Kepler Field

### 2024-05-21 paper/ms.tex
update ms

\title{Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field}

@@

\author{and coauthors, including but not limited to...}

@@

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous work; lithium is
  measured from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which our method is
  applicable.
  %
  We find that our rotation-based ages accurately recover ages of
  stars in open clusters; they also agree with
  $\gtrsim$90\% of the independent lithium ages.
  %
  The results include \nplyounggyrotwosigma\ planets
  younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
  at 3$\sigma$).
  %
  More broadly, we find that the star formation rate in the Kepler field
  has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over the past
  3\,Gyr.  This trend holds for both planet hosts and for the overall
  stellar sample.
  %
  This ``demographic cliff'' is a known feature of the Galaxy's star
  formation history, and it helps clarify the relative balance of
  young and old planets that have been detected to date using
  transits.
  %
\end{abstract}

@@

Young ($<$1\,Gyr) exoplanets therefore represent one means of building
the timeline for exoplanet evolution.  Informative individual
exemplars include V1298~Tau, an adolescent resonant chain of four
transiting puffy planets that is a likely precusor to the compact
multiplanet systems \citep{David_2019}, and HIP~67522, a Jupiter-sized
planet with a Neptunian mass (\citealt{Rizzuto_2020}; Thao et
al.~submitted).  Population-level analyses have similarly suggested
differences in the size distribution of young exoplanets relative to
their older counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

Discovering a young planet requires solving two problems: find the
planet, and measure the star's age.  Each problem admits a range of
solutions
\citep[e.g.][]{2008Sci...322.1348M,2012ApJ...756L..33Q,2024AJ....167..193T}.
In this article we will consider planets whose existence is previously
known from transits, and infer new stellar ages using rotation and
lithium.

@@

Despite this and other observational challenges, young close-in planet
discovery has matured over the past decade, primarily due to Kepler,
K2, TESS, and Gaia
\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Barber_2022,Zhou_2022,Zakhozhay_2022,Wood_2023}.
The strategy pursued by most groups during the 2010s was to focus on
stars with known ages---obvious members of open clusters---and to
search them for transiting planets.  The resulting stellar (and
assumed planetary) ages are precise at the $\approx$10\% level.  A
more recent development, facilitated by the second Gaia data release,
is the idea that finding transiting planets is often easier than
finding the host stellar association \citep[e.g.][]{Tofflemire_2021}.

The ``cluster-first'' approach has a major limitation: most nearby
sub-gigayear stars are in the field.  Only $\lesssim$1\% of stars
within $\approx$500\,pc have been associated with their birth cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
One can appreciate the problem by querying the NASA Exoplanet Archive
\citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
than 1\,Gyr.  Imposing a precision requirement of say $t$$<$1\,Gyr at
2$\sigma$ precision gives $\approx$50 ``young'' transiting planets at
the time of writing.  Most of these young planets are in clusters, and
were found by Kepler, K2, or TESS.  However, these surveys have
discovered $\approx$5{,}000 planets in total.  Assuming a constant
star formation history, we would expect an order of magnitude more
sub-gigayear transiting planets.

This study aims to resolve two main questions.  First, how wrong is it
to assume a uniform age distribution for transiting planet host stars?
Second, where are the missing young planets?  We will find that
uniform is wrong by a factor of a few, and that stellar activity may
be a less significant bias than the challenge of precisely determining
ages for field stars.

Despite Kepler's main mission ending quite some time ago, the ages of
many Kepler planets remain uncertain.  While isochrone ages have been
calculated for Kepler stars \citep{Berger_2020b_rpage} and Kepler
Objects of Interest \citep[KOIs;][]{Petigura_2022}, isochrone ages are
most precise for stars whose luminosities and temperatures separate
them from the main sequence.  For $\lesssim$1\,Gyr stars, isochrone
ages are therefore limited.

Stellar rotation periods offer a promising alternative.  The idea of
using a star's spin-down as a clock has a rich history
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015}.
Modern models yield ages precise to $\lesssim$30\% for FGK stars
between 1-4\,Gyr, and constraining age posteriors at earlier times
\citep{Bouma_2023}.  Physics-based models
\citep{Matt_2015,Gallet_Bouvier_2015,Spada_2020} are beginning to
connect these empirical relations to the evolution of stellar winds,
internal structure, and angular momentum transport.

In the context of Kepler, rotation-based ages have been reported for
planet-hosting stars since the early data releases
\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  Recent studies
focused on rotation-based ages for Kepler have further explored
incorporating information from stellar kinematics
\citep{2021AJ....161..189L,2024AJ....167..159L}, and from light curve
features such as starspot amplitudes \citep{2023ApJ...952..131M}.  Our
analysis is motivated by a few factors that we believe constitute
improvements, particularly for the ages of sub-gigayear stars.   These
factors are as follows.
\begin{enumerate}[label={\it \roman*)},leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The Kepler Object of Interest catalog, and our vetting of
    false positives within it, has reached maturity
    \citep[e.g.][]{Thompson_2018};
  %
  \item Measured rotation periods of FGK stars in open clusters now
    show not only the average evolution of $P_{\rm rot}(T_{\rm eff},t)$,
    but also how the astrophysical dispersion of stars around this average
    converges by the $\approx$700\,Myr age of Praesepe
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item We can fit the open cluster data, and use these fits to infer
    ages by marginalizing over the range of ages that might explain
    any one star's rotation period \citep{Bouma_2023}.
    For Sun-like stars, this model is consistent with older
    calibrations \citep[e.g.][]{Mamajek_2008}, though with an improved treatment
    of uncertainty propagation.  For late G dwarfs and
    K dwarfs, the calibration differs by factors of a few;
  %
  \item We can identify binary stars at greater fidelity today than
    in the past, which can clarify otherwise problematic estimates of
    rotation-based ages.
  %
  \item New open clusters in various stages of dissolution have been
    found in the Kepler field
    \cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,Barber_2022}.
    These clusters offer a sample of stars for which we can test the
    reliability of single-star age-dating methods.
\end{enumerate}

@@

We discuss our method for selecting the star and planet samples in
Section~\ref{sec:selection}, and describe the origin of our adopted
stellar parameters other than ages in Section~\ref{sec:stellarprops}.
We describe our age-dating methods in
Sections~\ref{sec:rotage} and~\ref{sec:liage}, and test them using
stars in clusters in the Kepler field.  The population-level
trends for both stars and planets are described in
Sections~\ref{sec:results} and~\ref{sec:disc}, and a few conclusions
are offered in Section~\ref{sec:conclusions}.

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
		}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf The stars.}  Our analysis focuses on stars observed by Kepler
    with previously reported rotation periods (dark gray points).  The
    rotation periods were primarily measured by
    \citet{Santos_2019,Santos_2021}.  About half of these rotators are
    suitable for gyrochronology (black points), based on factors
    including their temperatures, surface gravities, non-binarity, and
    proximity to the main sequence (see Section~\ref{subsec:flags}).
    Some are ``confirmed'' or ``candidate'' Kepler Objects of Interest
    that meet an additional set of planetary quality criteria (blue
    points; Section~\ref{subsec:plflags}).  Surface gravities and
    effective temperatures were derived photometrically by
    \citet{Berger_2020a_catalog}.
	}
	\label{fig:stellarprops}
\end{figure*}

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
		}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Rotation periods for Kepler target stars (left) and known
    planet hosts (right)}.  
    The gray lines are ``mean fits'' to the slow rotation sequences of
    open clusters.  The stellar sample (left) includes only apparently
    single stars near the main sequence with $\log g$$>$4.2,
    RUWE$<$1.4, and temperatures of 3800--6200\,K.  The color map has
    a linear stretch.  The planet-hosts (right) require the same
    stellar cuts, and include only the confirmed and candidate planets
    described in Section~\ref{subsec:plflags}.  An order of magnitude
    estimate for the number of young planets discovered by Kepler
    follows by counting planets below the mean rotational isochrones.
	}
	\label{fig:prot_vs_teff}
\end{figure*}

@@

We measured the lithium equivalent widths using a procedure adapted
from previous work \citep{Bouma_2021}.  Our stars of interest are FGK
stars, and so the continuum in the vicinity of the \ion{Li}{1}
6708\,\AA\ doublet is well-defined.  We Doppler-corrected the spectra
to a common reference wavelength by cross-correlating against a high
S/N template for Kepler-1698, chosen because $|\gamma|$$<$10\kms,
$T_{\rm eff}$$\approx$5000\,K and $v\sin i$$\approx$5\,\kms, which
puts it in the middle of our sample's temperature range and gives it
mild line-broadening.  We then trimmed the Doppler-corrected spectra
to a local window centered on the lithium line, using a window width
of 15\,\AA\ (we also considered 10\,\AA\ and 20\,\AA; the results were
consistent).  We continuum-normalized by fitting a third-degree
Chebyshev series, while excluding regions with absorption lines.  We
then numerically integrated the resulting spectrum using a
one-component Gaussian with free amplitude, width, and mean, and
estimated uncertainties on the line width through a Monte Carlo
procedure that bootstrapped against the local scatter in the continuum.

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.47\textwidth]{li_vs_teff_koi_X_JUMP_eagles.pdf}
	\end{center}
	\vspace{-0.6cm}
	\caption{
    {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
    for planet-hosting stars.} These measurements were made from
    Keck/HIRES spectra collected between 2009 and 2024.  Lines are the ``mean''
    isochrones from the same models as in Figure~\ref{fig:models}.  The
    intrinsic dispersion at $\gtrsim$1\,Gyr becomes much larger
    than changes in the mean.  Some stars with lithium detections do
    not have detected rotation periods, and vice-versa.
		\label{fig:li_vs_teff}
	}
\end{figure}

@@

Our default source for stellar temperatures and surface gravities is
the Gaia-Kepler Stellar Properties Catalog (GKSP;
\citealt{Berger_2020a_catalog}).  The GKSP parameters were reported
for stars with ``AAA'' 2MASS photometry, measured parallaxes in Gaia
DR2,  and $g$-band photometry available from either SDSS or else Kepler-INT
\citep{2012AJ....144...24G}.  The parameters themselves were derived using
\texttt{isoclassify} \citep{2017ApJ...844..102H} to interpolate over
the MIST isochrone grids \citep{Choi_2016,2016ApJS..222....8D}, given
the SDSS $g$ and 2MASS $K_{\rm s}$ photometry, the Gaia DR2
parallaxes, and $V$-band extinction from the
\citet{2018MNRAS.478..651G} reddening map.  The resulting stellar
parameters are available for \fracstarswithprotwithbtwenty\ of the
\nuniqstarsantosrot\ Kepler stars with rotation periods.  For the
remaining \fracstarswithprotwithoutbtwenty\ of stars that lack
temperatures and surface gravities from
\citetalias{Berger_2020a_catalog}, we adopt the values reported by
\citet{Santos_2019} and \citet{Santos_2021}, which for these cases
primarily derive from the \citet{Mathur_2017} DR25 Kepler Stellar
Properties Catalog, and are mostly derived from photometry.  In the
planet sample, \frackoisnofpwithprotwithbtwenty\ of the
non-false-positive KOIs with rotation periods have parameters from
\citet{Berger_2020a_catalog}, and the remainder similarly are drawn
from DR25.

\citet{David_2021} compared the photometric
\citetalias{Berger_2020a_catalog} stellar parameters ($T_{\rm eff}$,
$R_\star$, [Fe/H]) against the spectroscopic parameters from
\citet{Fulton_2018}.  The temperature scales did show a few-percent
systematic difference, with \citetalias{Fulton_2018} quoting higher
temperatures than \citetalias{Berger_2020a_catalog} for mid K dwarfs,
and lower temperatures for early F dwarfs.  Our age analysis,
described below, assumes Gaussian uncertainties in effective
temperatures; we by default adopt the maximum of the two-sided
uncertainties reported by \citetalias{Berger_2020a_catalog} as our
symmetric temperature uncertainty.  Overall, the impact of systematic
uncertainties in the temperature scale on our age uncertainties is
generally smaller than that of the intrinsic scatter in the
calibration sequences of the open clusters.

@@

We calculate gyrochrone ages using \texttt{gyro-interp}
\citep{Bouma_2023}, which is a method designed to address the fact
that stars with the same mass and same rotation period can have very
different ages \citep[e.g.][]{Curtis_2019_ngc6811}.  The gyrochrone
age posterior should therefore incorporate this intrinsic
population-level scatter into the precision with which the age can be
inferred.  Figure~\ref{fig:models} highlights this problem,
particularly in regions where the 0.1-1\,Gyr stars overlap.

@@

\subsection{Stellar Quality Flags}
\label{subsec:flags}
We calculated gyrochrone ages for all \nuniqstarsantosrotteffcut\
stars with reported rotation periods and effective temperatures of
3800--6200\,K.  To flag stars for which we suspect these ages may not
in fact be valid, we built a set of quality flags, which we condensed
into a single binary number, $Q_{\rm star}$.  When and how the flags
represented in this should be applied depends on the question being
asked.  If the goal is to construct a false-positive free sample, they
could all be applied.  If the goal is to construct a complete sample,
then consider the examples of Kepler~1627Ab ($\approx$40\,Myr) and
Kepler~51c ($\approx$625\,Myr).  The former has a high RUWE due to a
resolved binary companion \citep{Bouma_2022a}; the latter is on a
grazing orbit \citep{2014ApJ...783...53M}.  We leave selection for or
against such cases as a decision to the user.

@@

{\it Kepler-Gaia crossmatch quality (bits 4 and 5)}--- In situations
where we wanted to leverage Gaia DR3 data, we used M.~Bedell's 4$''$
Kepler-to-Gaia crossmatch %\footnote{\url{https://gaia-kepler.fun}, last
%accessed 15 April 2024.  We used 4$''$ rather than the 1$''$ match
%because 91 KOIs are omitted in the 1$''$ match, 23 of which have
%either ``candidate'' or ``confirmed'' status.  Almost all of these 23
%are M dwarfs with high proper motions.  Performing the same crossmatch
%at 4$''$, only two KOIs (KOI-3993 and KOI-6531), both known
%false positives, fail to yield Gaia DR3 crossmatches.}, 
%which is a
%crossmatch
of the NEA \texttt{q1\_q17\_dr25\_stellar} catalog with Gaia DR3 (accessible
online at \url{https://gaia-kepler.fun}).  The
separation distribution of the Kepler-Gaia DR3 crossmatches is such
that 99.2\% of candidate matches are within 1$''$.   We nonetheless
noted an upturn in the candidate match rate from 3-4$''$; such sources
are flagged using bit 4.  For KIC stars with multiple potential Gaia
matches within the 4$''$ radius, we adopted the brightest star as the
default match.  In most such cases this was unambiguous because there
is a large brightness difference between the primary and any apparent
neighbors.  However cases with multiple stars within 4$''$ within
$\Delta G$$<$$0.5$\,mag are noted using bit flag 5.

@@

Stars that formed in the same birth cluster 
are the gold standard for the astronomical age scale
\citep{Soderblom_2010}.
Before Gaia, the few $\gtrsim$1\,Gyr open clusters
known in the Kepler field had been visually identified by
\citet{1864RSPT..154....1H}.
Gaia proper motions and distances 
enabled the discovery of stellar ensembles that are more diffuse,
but which nonetheless
share a common age 
based on e.g.~isochrone, rotation, and lithium dating
\cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,Barber_2022}.

>>>>

%\title{Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field}
\title{A Dearth of Young Stars and Planets in the Kepler Field}

@@

%\author{and coauthors, including but not limited to...}

@@

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous work; lithium is
  measured from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which our method is
  applicable.
  %
  We find that our rotation-based ages accurately recover ages of
  stars in open clusters spanning 0.04-2.5\,Gyr; they also agree with
  $\gtrsim$90\% of the independent lithium ages.
  %
  The results include \nplyounggyrotwosigma\ planets
  younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
  at 3$\sigma$).
  %
  This is far fewer than expected under the classic assumption of a uniform
  star formation history: the star formation rate in the Kepler field
  has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over the past
  3\,Gyr.  This trend holds for both planet hosts and for the overall
  stellar sample.
  %
  This ``demographic cliff'' in the Galaxy's star
  formation history has been previously reported, and it helps clarify the relative balance of
  young and old planets that have been detected to date using
  transits.
  %
\end{abstract}

@@

Young ($<$1\,Gyr) exoplanets therefore represent one means of building
the timeline for exoplanet evolution.  Informative individual
exemplars include V1298~Tau, a resonant chain of 
puffy planets that is a likely precusor to the compact
multiplanet systems \citep{David_2019}, and HIP~67522, a Jupiter-sized
planet with a Neptunian mass (\citealt{Rizzuto_2020}; Thao et
al.~submitted).  Population-level analyses have similarly suggested
differences in the size distribution of young exoplanets relative to
their older counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

Discovering a young planet requires solving two problems: find the
planet, and measure the star's age.  Each problem admits a range of
solutions
\citep[e.g.][]{2008Sci...322.1348M,2012ApJ...756L..33Q,2024AJ....167..193T}.
In this article we will consider planets whose existence has been
previously established using transits, and infer new stellar ages
using rotation and lithium.

@@

Despite this and other observational challenges, young close-in planet
discovery has matured over the past decade, primarily due to Kepler,
K2, TESS, and Gaia
\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Barber_2022,Zhou_2022,Zakhozhay_2022,Wood_2023}.
The strategy pursued by most groups during the 2010s was to focus on
stars with known ages---obvious members of open clusters---and to
search them for transiting planets.  The resulting stellar (and
assumed planetary) ages are precise at the $\approx$10\% level.  A
more recent development, facilitated by Gaia, is the idea that finding
transiting planets is often easier than finding the host stellar
association \citep[e.g.][]{Tofflemire_2021}.

The ``cluster-first'' approach has a major limitation: most nearby
sub-gigayear stars are in the field.  Only $\lesssim$1\% of stars
within $\approx$500\,pc have been associated with their birth cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
One can appreciate the problem by querying the NASA Exoplanet Archive
\citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
than 1\,Gyr.  Requiring say $t$$<$1\,Gyr at
2$\sigma$ precision gives $\approx$50 ``young'' transiting planets at
the time of writing.  Most of these young planets are in clusters, and
were found by Kepler, K2, or TESS.  However, these surveys have
discovered $\approx$5{,}000 planets in total.  Assuming a constant
star formation history, we would expect an order of magnitude more
sub-gigayear transiting planets.

This study aims to resolve two questions.  First, how wrong is it
to assume a uniform age distribution for transiting planet host stars?
Second, where are the missing young planets?  We will find that
uniform is wrong by a factor of a few, and that stellar activity may
be a less significant bias in young transiting planet detection than
the challenge of precisely determining ages for field stars.

Despite Kepler's main mission ending quite some time ago, the ages of
many Kepler planets remain uncertain.  While isochrone ages have been
calculated for Kepler stars \citep{Berger_2020b_rpage} and Kepler
Objects of Interest \citep[KOIs;][]{Petigura_2022}, such ages are
most precise for stars whose luminosities and temperatures separate
them from the main sequence.  For $\lesssim$1\,Gyr stars, isochrone
ages are therefore limited.

Stellar rotation periods offer a promising alternative.  The idea of
using a star's spin-down as a clock has a rich history
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015}.
Empirical models now yield ages precise to $\lesssim$30\% for FGK stars
between 1-4\,Gyr, and constraining age posteriors at earlier times
\citep{Bouma_2023}.  Physics-based models
\citep{Matt_2015,Gallet_Bouvier_2015,Spada_2020} can 
connect these empirical relations to the evolution of stellar winds,
internal structure, and angular momentum transport.

Rotation-based ages have been reported for
various subsets of Kepler stars since the early data releases
\citep[e.g.][]{Walkowicz_2013,McQuillan_2014,Reinhold_2015,Angus_2018}.
More recent work has further explored incorporating information from
stellar kinematics \citep{2021AJ....161..189L,2024AJ....167..159L},
and from light curve features such as starspot amplitudes
\citep{2023ApJ...952..131M}.  Our analysis is motivated by a few
factors that can yield improvements, particularly for the ages of
sub-gigayear stars.   These factors are as follows.
\begin{enumerate}[label={\it \roman*)},leftmargin=12pt,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The Kepler Object of Interest catalog, and our vetting of
    false positives within it, has now reached maturity
    \citep[e.g.][]{Thompson_2018};
  %
  \item Measured rotation periods of FGK stars in open clusters now
    show not only the average evolution of $P_{\rm rot}(T_{\rm eff},t)$,
    but also how the astrophysical dispersion of stars around this average
    converges by the $\approx$700\,Myr age of Praesepe
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item We can fit the open cluster data, and use these fits to infer
    ages by marginalizing over the range of ages that might explain
    any one star's rotation period \citep{Bouma_2023}.
    This represents an improvement in both accuracy and in realism
    of uncertainty propagation relative to previous calibrations.
  %
  \item We can identify binary stars with greater fidelity than
    in the past, which can clarify otherwise problematic estimates of
    rotation-based ages.
  %
  \item New open clusters in various stages of dissolution have been
    found in the Kepler field
    \cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,Barber_2022}.
    These clusters offer a sample of stars with which we can test the
    reliability of single-star age-dating methods.
\end{enumerate}

@@

We discuss our method for selecting the star and planet samples in
Section~\ref{sec:selection}, and describe the origin of our adopted
stellar parameters other than ages in Section~\ref{sec:stellarprops}.
We describe our age-dating methods in
Sections~\ref{sec:rotage} and~\ref{sec:liage}, and test them in
Section~\ref{sec:litagecomparison} using
clusters in the Kepler field.  The population-level
trends are described in
Sections~\ref{sec:results} and~\ref{sec:disc}, and a few conclusions
are offered in Section~\ref{sec:conclusions}.

@@

\begin{figure}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
		}

		\subfloat{
			\includegraphics[width=0.48\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf The stars.}  Our analysis focuses on stars observed by Kepler
    with previously reported rotation periods (dark gray points).  The
    rotation periods are primarily drawn from
    \citet{Santos_2019,Santos_2021}.  About half of these rotators are
    suitable for gyrochronology (black points), based on factors
    including their temperatures, surface gravities, non-binarity, and
    proximity to the main sequence (see Section~\ref{subsec:flags}).
    Some are ``confirmed'' or ``candidate'' Kepler Objects of Interest
    that meet an additional set of planetary quality criteria (blue
    points; Section~\ref{subsec:plflags}).  Surface gravities and
    effective temperatures were derived photometrically by
    \citet{Berger_2020a_catalog}.
    MIST isochrones spanning 3--6\,Gyr in 1\,Gyr increments 
    are shown in the lower panel \citep{Choi_2016} .
	}
	\label{fig:stellarprops}
\end{figure}

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
		}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Rotation periods for Kepler target stars (left) and known
    planet hosts (right)}.  
    The gray lines are ``mean fits'' to the slow rotation sequences of
    open clusters.  The stellar sample (left) includes only apparently
    single stars near the main sequence with $\log g$$>$4.2,
    RUWE$<$1.4, and temperatures of 3800--6200\,K.  The color map has
    a linear stretch.  The planet-hosts (right) require the same
    stellar cuts, and include only the confirmed and candidate planets
    described in Section~\ref{subsec:plflags}.  An order of magnitude
    estimate for the number of young planets discovered by Kepler
    follows by counting planets below the mean rotational isochrones.
	}
	\label{fig:prot_vs_teff}
\end{figure*}

@@

We measured the lithium equivalent widths using a procedure adapted
from previous work \citep{Bouma_2021}.  Our stars of interest are FGK
stars, and so the continuum in the vicinity of the \ion{Li}{1}
6708\,\AA\ doublet is well-defined.  We Doppler-corrected the spectra
to a common reference wavelength by cross-correlating against a high
S/N template for Kepler-1698, chosen because $|\gamma|$$<$10\kms,
$T_{\rm eff}$$\approx$5000\,K and $v\sin i$$\approx$5\,\kms, which
puts it in the middle of our sample's temperature range and gives it
mild line-broadening.  We then trimmed the Doppler-corrected spectra
to a local window centered on the lithium line, using a window width
of 15\,\AA\ (we also considered 10\,\AA\ and 20\,\AA; the results were
consistent).  We continuum-normalized by fitting a third-degree
Chebyshev series, while excluding regions with absorption lines.  We
then numerically integrated the resulting spectrum using a
one-component Gaussian with free amplitude, width, and mean, and
estimated uncertainties on the line width through a Monte Carlo
procedure that bootstrapped against the local scatter in the continuum.
The resulting EWs are shown in Figure~\ref{fig:li_vs_teff}.

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.48\textwidth]{li_vs_teff_koi_X_JUMP_eagles.pdf}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
    for planet-hosting stars.} These measurements were made from
    Keck/HIRES spectra collected from 2009--2024.  Lines are the ``mean''
    isochrones from the same models as in Figure~\ref{fig:models}.  The
    intrinsic dispersion becomes much larger
    than changes in the mean at $\gtrsim$1\,Gyr.  Some stars with lithium detections do
    not have detected rotation periods, and vice-versa.
		\label{fig:li_vs_teff}
	}
\end{figure}

@@

Our default source for stellar temperatures and surface gravities is
the Gaia-Kepler Stellar Properties Catalog (GKSP;
\citealt{Berger_2020a_catalog}).  The GKSP parameters were reported
for stars with ``AAA'' 2MASS photometry, measured parallaxes in Gaia
DR2,  and $g$-band photometry available from either SDSS or else Kepler-INT
\citep{2012AJ....144...24G}.  The parameters themselves were derived using
\texttt{isoclassify} \citep{2017ApJ...844..102H} to interpolate over
the MIST isochrone grids \citep{Choi_2016,2016ApJS..222....8D}, given
the SDSS $g$ and 2MASS $K_{\rm s}$ photometry, the Gaia DR2
parallaxes, and $V$-band extinction from the
\citet{2019ApJ...887...93G} reddening map.  The resulting stellar
parameters are available for \fracstarswithprotwithbtwenty\ of the
\nuniqstarsantosrot\ Kepler stars with rotation periods.  For the
remaining \fracstarswithprotwithoutbtwenty\ of stars that lack
temperatures and surface gravities from
\citetalias{Berger_2020a_catalog}, we adopt the values reported by
\citet{Santos_2019} and \citet{Santos_2021}, which for these cases
primarily derive from the \citet{Mathur_2017} DR25 Kepler Stellar
Properties Catalog, and are mostly derived from photometry.  In the
planet sample, \frackoisnofpwithprotwithbtwenty\ of the
non-false-positive KOIs with rotation periods have parameters from
\citet{Berger_2020a_catalog}, and the remainder similarly are drawn
from DR25.

\citet{David_2021} compared the photometric
\citetalias{Berger_2020a_catalog} stellar parameters ($T_{\rm eff}$,
$R_\star$, [Fe/H]) against the spectroscopic parameters from
\citet{Fulton_2018}.  The temperature scales showed a few-percent
systematic difference, with \citetalias{Fulton_2018} quoting higher
temperatures than \citetalias{Berger_2020a_catalog} for mid K dwarfs,
and lower temperatures for early F dwarfs.  Our age analysis,
described below, adopts the maximum of the two-sided
uncertainties reported by \citetalias{Berger_2020a_catalog} as a
symmetric Gaussian temperature uncertainty.  Systematic
uncertainties in the temperature scale generally influence 
our age uncertainties at a
smaller level than the statistical intrinsic scatter in the
open cluster rotation sequences.

@@

We calculate gyrochrone ages using \texttt{gyro-interp}
\citep{Bouma_2023}, which is a method designed to address the fact
that stars with the same mass and same rotation period can have very
different ages \citep[e.g.][]{Curtis_2019_ngc6811}.  The gyrochrone
age posterior should therefore incorporate this intrinsic
population-level scatter into the precision of its reported
ages.  Figure~\ref{fig:models} highlights this problem,
particularly in regions where the 0.1-1\,Gyr stars overlap.

@@

\subsection{Stellar Quality Flags}
\label{subsec:flags}
We calculated gyrochrone ages for all \nuniqstarsantosrotteffcut\
stars with reported rotation periods and effective temperatures of
3800--6200\,K.  To flag stars for which we suspect these ages may not
in fact be valid, we built a set of quality flags, which we condensed
into a single binary number, $Q_{\rm star}$.  When and how the flags
represented in this bitmask should be applied depends on the question being
asked.  If the goal is to construct a false-positive free sample, they
could all be applied.  If the goal is to construct a complete sample,
then consider the examples of Kepler~1627Ab ($\approx$40\,Myr) and
Kepler~51c ($\approx$625\,Myr).  The former has a high RUWE due to a
resolved binary companion \citep{Bouma_2022a}; the latter is on a
grazing orbit \citep{2014ApJ...783...53M}.  We leave selection for or
against such cases as a decision to the user.

@@

{\it Kepler-Gaia crossmatch quality (bits 4 and 5)}--- In situations
where we wanted to leverage Gaia DR3 data, we used M.~Bedell's 4$''$
Kepler-to-Gaia crossmatch %\footnote{\url{https://gaia-kepler.fun}, last
%accessed 15 April 2024.  We used 4$''$ rather than the 1$''$ match
%because 91 KOIs are omitted in the 1$''$ match, 23 of which have
%either ``candidate'' or ``confirmed'' status.  Almost all of these 23
%are M dwarfs with high proper motions.  Performing the same crossmatch
%at 4$''$, only two KOIs (KOI-3993 and KOI-6531), both known
%false positives, fail to yield Gaia DR3 crossmatches.}, 
%which is a
%crossmatch
of the NEA \texttt{q1\_q17\_dr25\_stellar} catalog with Gaia DR3
(available at \url{https://gaia-kepler.fun}).  The
separation distribution of the Kepler-Gaia DR3 crossmatches is such
that 99.2\% of candidate matches are within 1$''$.   We nonetheless
noted an upturn in the candidate match rate from 3-4$''$; such sources
are flagged using bit 4.  For KIC stars with multiple potential Gaia
matches within the 4$''$ radius, we adopted the brightest star as the
default match.  In most such cases this was unambiguous because there
is a large brightness difference between the primary and any apparent
neighbors.  However cases with multiple stars within 4$''$ within
$\Delta G$$<$$0.5$\,mag are noted using bit flag 5.

@@

Stars that formed in the same birth cluster 
are the gold standard for the astronomical age scale
\citep{Soderblom_2010}.
Before Gaia, the few $\gtrsim$1\,Gyr open clusters
known in the Kepler field had been visually identified by
\citet{1864RSPT..154....1H}.
Gaia proper motions and distances 
have enabled the discovery of stellar ensembles that are more diffuse,
but which nonetheless
share a common age 
based on e.g.~isochrone, rotation, and lithium dating
\cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,Barber_2022}.
